Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

\usepackage{makecell}

@@

The LS model is a convenient probabilistic approximation to the coalescent with
recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
as an HMM in which a focal genome is a sequence of nucleotides (observed
states) that are probabilistically emitted as an imperfect mosaic of a set of
reference genomes (hidden states). As such, an efficient HMM machinery can be
employed to make statements about parameters of interest and highlight
potential sources of error in genotyping and sequencing data. The LS model is
used in a wide variety of applications in genomics. For example, it is the
engine that drives statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
\citep{Hinch2011-tz}), and selection within and across hosts in viral sequence
data (e.g., \citep{Palmer2019-wa}).

@@

Both the wide and long ARGs contain a large number of recombination nodes
(4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of
these nodes inferred as a result of matching a specific sample (a recombinant)
into the ARG. However, ~65\% of recombination nodes are ancestors of only a
single descendant sample in each local tree; these could have arisen from
sequencing of mixed samples (e.g., from co-infection or contamination), or were
simply not transmitted onwards and so were excluded from subsequent analyses
(leaving X and Y recombinants in the wide and long ARGs, respectively). In
their analysis using RIPPLES, Turakhia et al. (2022) detected 589 recombinants,
also excluding singletons.

@@

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by Turakhia et al.
\citep{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
the number of intersecting intervals in the N gene following the peak in ORF8
towards the 3’ end.

@@

In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. Yi et al. \citep{Yi2021-sc} reconstructed a SARS-CoV-2
phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
January 12, 2021, and systematically catalogued all the mutations occurring
along the phylogeny. Notably, in the mutational spectrum described by Yi et
al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
mutations, respectively.

@@

Our results show that sc2ts identifies the early recombinants (pre-2022)
previously identified by Jackson et al. (2021). Sc2ts also identifies many
Pango recombinants, which are designated after community perusal of supporting
genomic evidence. We have found several previously unreported and overlooked
recombinant sequences in the wide ARG. After further improvements and
enhancements, we will use sc2ts to build a comprehensive ARG to search for
recombination signals among all the 15 million and counting SARS-CoV-2 genomes
available.

>>>>

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

The LS model is a convenient probabilistic approximation to the coalescent with
recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
as an HMM in which a focal genome is a sequence of nucleotides (observed
states) that are probabilistically emitted as an imperfect mosaic of a set of
reference genomes (hidden states). As such, an efficient HMM machinery can be
employed to make statements about parameters of interest and highlight
potential sources of error in genotyping and sequencing data. The LS model is
used in a wide variety of applications in genomics. For example, it is the
engine that drives statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral sequence
data \citep[e.g.][]{Palmer2019-wa}.

@@

Both the wide and long ARGs contain a large number of recombination nodes
(4,123 and 2,078, respectively; Supplementary Tables 1 and 2), with each of
these nodes inferred as a result of matching a specific sample (a recombinant)
into the ARG. However, ~65\% of recombination nodes are ancestors of only a
single descendant sample in each local tree; these could have arisen from
sequencing of mixed samples (e.g., from co-infection or contamination), or were
simply not transmitted onwards and so were excluded from subsequent analyses
(leaving X and Y recombinants in the wide and long ARGs, respectively). In
their analysis using RIPPLES, \cite{Turakhia2022-it} detected 589 recombinants,
also excluding singletons.

@@

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by
\cite{Turakhia2022-it}.
Inspecting more closely, we find an enrichment in the
number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
the number of intersecting intervals in the N gene following the peak in ORF8
towards the 3’ end.

@@

In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
January 12, 2021, and systematically catalogued all the mutations occurring
along the phylogeny. Notably, in the mutational spectrum described by Yi et
al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
mutations, respectively.

@@

Our results show that sc2ts identifies the early recombinants (pre-2022)
previously identified by \cite{Jackson2021-ik}. Sc2ts also identifies many
Pango recombinants, which are designated after community perusal of supporting
genomic evidence. We have found several previously unreported and overlooked
recombinant sequences in the wide ARG. After further improvements and
enhancements, we will use sc2ts to build a comprehensive ARG to search for
recombination signals among all the 15 million and counting SARS-CoV-2 genomes
available.
